---
title: "Pseudobulk"
subtitle: "Run reintegration after cleaning"
date: today
embed-resources: true
dpi: 300
execute: 
  warning: false
  echo: false
editor_options: 
  chunk_output_type: console
---

```{r}

library(tidyverse)
library(Seurat)
library(patchwork)
library(kableExtra)
library(jsonlite)
library(compareGroups)
library(EnhancedVolcano)
```

## REINTEGRATION

The integrated scRNAseq dataset of all brain cell populations was subsetted to retain those cells no subsetted out by the "expert committee".

```{r}
#| eval: false

# Load the full annotated Seurat object containing all brain cell populations
load("data/20250812-alldata_sctype_annotated_CHLOE_Lance.Rdata")
alldata_clean$original_clusters <- alldata_clean$seurat_clusters

# Calculate ribosomal gene percentage in the SCT assay
alldata_clean$ribo_percent <- PercentageFeatureSet(alldata_clean, assay = "SCT", pattern = "^Rp[ls]")
# Calculate mitochondrial gene percentage in the SCT assay
alldata_clean$mito_percent <- PercentageFeatureSet(alldata_clean,assay = "SCT", pattern = "^mt")
# Remove cells with mito percen > 30% (There is none)
alldata_clean <- subset(alldata_clean,subset = mito_percent<30)

```

```{r}
#| eval: false
### some cleaning
DefaultAssay(alldata_clean) <- "RNA"
alldata_clean@graphs <- list()
alldata_clean@reductions <- list()
alldata_clean[["integrated"]] <- NULL
alldata_clean[["SCT"]] <- NULL

### empty metadata columns with or with unique values
alldata_clean@meta.data$Notes...10 <- NULL
alldata_clean@meta.data$Notes...12 <- NULL
alldata_clean@meta.data$Strain <- NULL
alldata_clean@meta.data$Labels <- NULL
########################################################

# split by sample into a list
objs <- SplitObject(alldata_clean, split.by = "Mouse.ID")
# SCTransform
objs <- map(objs, \(x) SCTransform(x))



# Select features and prep for integration
features <- SelectIntegrationFeatures(objs, nfeatures = 2000)
objs <- PrepSCTIntegration(objs, anchor.features = features)

# Find anchors and integrate
anchors <- FindIntegrationAnchors(object.list = objs,
                                  normalization.method = "SCT",
                                  reduction = "cca",
                                  anchor.features = features)

aria <- IntegrateData(anchorset = anchors, normalization.method = "SCT")

# Run PCA on variable features from the integrated assay
aria <- RunPCA(
  aria,
  assay   = "integrated",
  verbose = FALSE
)

# Set seed for reproducibility
set.seed(1234)

# Compute nearest neighbors using first 10 PCs
aria <- FindNeighbors(aria, reduction = "pca", dims = 1:30)

# Perform clustering at multiple resolutions for later comparison
aria <- FindClusters(aria, resolution = 0.8)

# Select the resolution of interest (0.3) and store as "final_clusters"
aria$final_clusters_afterReintegration <- aria$integrated_snn_res.0.8
Idents(aria) <- "final_clusters_afterReintegration"



aria <- RunUMAP(aria, reduction = "pca", dims = 1:30, verbose = FALSE)

# save(aria, file = "data/20250826-alldata_sctype_annotated_CHLOE_Lance_reintegrated.Rdata")

```

## PCA

```{r}
#| column: page
#| fig-width: 12

load("data/20250826-alldata_sctype_annotated_CHLOE_Lance_reintegrated.Rdata")

pca <- PCAPlot(aria)
elbow <- ElbowPlot(aria, ndims = 50)

pca|elbow
```

<details>

<summary>__PC details__</summary>

```{r}
#| fig-width: 12
#| fig-height: 10

VizDimLoadings(aria, dims = 1:3)
```

1.  PC1 captures a strong myelination axis, with high loadings for oligodendrocyte and myelin-related genes (*Plp1, Mbp, Mag, Mal, Car2, Cryab*) and opposite contributions from astrocytic (*Ptgds, Enpp2, Edil3*) and microglial (*Hexb, C1qa, Ctss, Cst3*) markers.

2.  PC2 reflects a neuron–astrocyte interaction gradient, driven by the astrocytic ECM regulator *Sparcl1* on one pole and the neuronal signaling gene *Plcb1* on the other.

3.  PC3 highlights synaptic connectivity and neurotransmission, with neuronal adhesion and plasticity genes (*Nrxn1, Ntm, Lsamp, Ctnnd2, Adgrb3, Ntrk2*) and the inhibitory receptor *Gabrb1* loading strongly, balanced by astrocytic transporters (*Slc1a2, Slc4a4*) and *Ptprz1*. Together, these axes describe complementary layers of variance corresponding to myelination, glial–neuronal signaling, and synaptic specialization.

</details>
```{r}
#| eval: false
### highest loading in the first 3 PCs
PC1 <- TopFeatures(aria,nfeatures = 10, dim = 1)
PC2 <- TopFeatures(aria,nfeatures = 10, dim = 2)
PC3 <- TopFeatures(aria,nfeatures = 10, dim = 3)

hpc <- tibble(PC1, PC2, PC3)

kbl(hpc) |> 
    kable_styling("striped")

```

## UMAP after reintegration

```{r}
#| column: page
#| fig-width: 12




Idents(aria) <- "final_clusters_afterReintegration"
d1 <- DimPlot(aria)

Idents(aria) <- "lance_classification"
d2 <- DimPlot(aria)+
    scale_colour_brewer(palette = "Paired")

d1|d2

```

```{r}
source("20250725-helper_functions.R")
## scProp_help()
ag <- scProp(aria,
             cluster_col = "lance_classification",
             sample_col = "Mouse.ID",
             treatment_col = "Treatment.Group")
```

```{r}
ag$counts_plot
```

```{r}
#| fig-width: 10
#| fig-height: 4
ag$proportions_plot
```

```{r}
export2md(ag$summary_table, caption = "")
```


## Pseudobulk

```{r}

pseudo <- AggregateExpression(aria, assays = "RNA",
                              group.by = c("Mouse.ID","lance_classification","Treatment.Group"),
                              return.seurat = T)

pseudo$pseudobulk_id = paste(pseudo$lance_classification , pseudo$Treatment.Group,sep="_")

Idents(pseudo) <- "pseudobulk_id"
```

### Microglia

```{r}
### for help with the anayet (volcano) function, run anayet_help()

micro <- anayet(pseudo,
       ident1 = "Microglia_Adu",
       ident2 = "Microglia_IgG")

kbl(micro$significant_results) |> 
    kable_styling("striped")
```
Following treatment, microglia from the Alzheimer’s disease mouse model exhibited a broad transcriptional shift consistent with reduced pro-inflammatory activation and metabolic demand. Several mitochondrial genes (*mt-Nd4l*, *mt-Atp8*) and the amino acid transporter *Slc38a2* were downregulated, suggesting diminished oxidative phosphorylation and nutrient uptake that normally support highly activated microglial states. Key inflammatory mediators, including *Il1a*, *Cxcl1*, *Tnfaip2*, and the microRNA host gene *Mir155hg*, also decreased, indicating attenuation of cytokine release, chemotactic signaling, NF-κB pathway activity, and miR-155–driven immune activation. Concomitant reductions in *Prdm1* and regulatory non-coding transcripts (*Gm13849*, *C430049B03Rik*), together with decreased *Rpl35* and *Man1a*, point to lower transcriptional activity, protein synthesis, and secretory function, while the decline in *H2ac20* implies chromatin remodeling toward a less plastic state. Notably, this general dampening of inflammatory pathways was accompanied by increased expression of *Ccl7*, a chemokine involved in monocyte recruitment. This suggests that while most pro-inflammatory signals were suppressed, the treatment selectively maintained or enhanced pathways favoring controlled immune cell trafficking, potentially reflecting a rebalancing of microglial responses rather than their complete silencing. Collectively, these findings support the notion that the treatment shifts microglia away from a broadly activated phenotype toward a more homeostatic yet immunologically competent state, which may contribute to mitigation of chronic neuroinflammation in Alzheimer’s disease.

### Astrocytes


```{r}
### for help with the anayet (volcano) function, run anayet_help()

astro <- anayet(pseudo,
       ident1 = "Astrocytes_Adu",
       ident2 = "Astrocytes_IgG")

kbl(astro$significant_results) |> 
    kable_styling("striped")
```

### BAM

```{r}
### for help with the anayet (volcano) function, run anayet_help()

bam <- anayet(pseudo,
       ident1 = "BAM_Adu",
       ident2 = "BAM_IgG")

kbl(bam$significant_results) |> 
    kable_styling("striped")
```

### Choroid

```{r}
### for help with the anayet (volcano) function, run anayet_help()

cho <- anayet(pseudo,
       ident1 = "Choroid_Adu",
       ident2 = "Choroid_IgG")

kbl(cho$significant_results) |> 
    kable_styling("striped")
```

### Endothelial

```{r}
### for help with the anayet (volcano) function, run anayet_help()

endo <- anayet(pseudo,
       ident1 = "Endothelial_Adu",
       ident2 = "Endothelial_IgG")


kbl(endo$significant_results) |> 
    kable_styling("striped")
```


### Ependymal

```{r}
### for help with the anayet (volcano) function, run anayet_help()

epe <- anayet(pseudo,
       ident1 = "Ependymal_Adu",
       ident2 = "Ependymal_IgG")


kbl(epe$significant_results) |> 
    kable_styling("striped")
```


### Neurons

```{r}


neu <- anayet(pseudo,
       ident1 = "Neurons_Adu",
       ident2 = "Neurons_IgG")


kbl(neu$significant_results) |> 
    kable_styling("striped")
```

### NPC

```{r}


npc <- anayet(pseudo,
       ident1 = "NPC_Adu",
       ident2 = "NPC_IgG")


kbl(npc$significant_results) |> 
    kable_styling("striped")
```


### Oligodendrocytes

```{r}


oli <- anayet(pseudo,
       ident1 = "Oligodendrocytes_Adu",
       ident2 = "Oligodendrocytes_IgG")


kbl(oli$significant_results) |> 
    kable_styling("striped")
```


### OPC

```{r}


opc <- anayet(pseudo,
       ident1 = "OPC_Adu",
       ident2 = "OPC_IgG")


kbl(opc$significant_results) |> 
    kable_styling("striped")
```

### T-cell

```{r}


tcell <- anayet(pseudo,
       ident1 = "T-cell_Adu",
       ident2 = "T-cell_IgG")


kbl(tcell$significant_results) |> 
    kable_styling("striped")
```

### VMC

```{r}


vmc <- anayet(pseudo,
       ident1 = "VMC_Adu",
       ident2 = "VMC_IgG")


kbl(vmc$significant_results) |> 
    kable_styling("striped")
```


### Number of DEGs up/down-regululated by aducanumab

```{r}
updown <- bind_rows(c(cell = "Microglia", micro$significant_results),
                c(cell = "Oligodendrocytes", oli$significant_results),
                c(cell = "Endothelial", endo$significant_results),
                c(cell = "Astrocytes", astro$significant_results),
                c(cell = "T-cell", tcell$significant_results)) |> 
    mutate(Regulation = ifelse(avg_log2FC>0, "Up","Down")) |> 
    group_by(cell, Regulation) |> 
    summarize(n=n())


ggplot(updown, aes(x =cell, y=n, fill = Regulation))+
    geom_col()+
    labs(y = "Number of DEGs", x="")+
    scale_fill_manual(values = c("blue","Red"))+
    theme_minimal()+
        theme(axis.title = element_text(size=16),
              axis.text.y = element_text(size=14),
              axis.text.x = element_text(size = 14, angle = 25, hjust = 1, vjust = 1))
```

