---
title: "SCTransform & integration"
date: today
embed-resources: true
execute: 
  warning: false
editor_options: 
  chunk_output_type: console
---

```{r}
suppressPackageStartupMessages({
    library(tidyverse)
    library(Seurat)
    # BiocManager::install('glmGamPoi') 
    library(glmGamPoi)
    library(future)
    library(compareGroups)
    })


### helper functions
source("20250725-helper_functions.R")
```


## Subset cells with nFeat_RNA > 2000

```{r}

load ("data/20200620-list_of_Seurat_objects.Rdata")

seurat_objects <- map(seurat_objects, \(x) subset(x, subset = nFeature_RNA > 2000))

```

```{r}

features <- imap (seurat_objects, \(x, name) tibble(seurat_object = name,
                                             n_features = x$nFeature_RNA,
                                             ribo = PercentageFeatureSet(x, pattern = "^Rp[ls]"))) |> 
    list_rbind()

ggplot(features,aes(n_features,col=seurat_object))+
    labs(y="", title = "Number of features /cell after subsetting")+
    scale_x_continuous( " ", breaks = seq(0,12000,1000), 
                       labels = scales::label_number(scale = 1/1000, suffix = "K"))+
    density_decor()

ggplot(features,aes(ribo,col=seurat_object))+
    labs(y="", title = "ribo content /cell after subsetting")+
    scale_x_continuous( " ", limits = c(0,30),
                        labels = scales::label_number(scale = 1/1, suffix = "%"))+
    density_decor()
   

```
---

__Number of cells (N) and features per cell (median [IQR]) in the subsetted dataset__

```{r}
#| column: page

c1 <- compareGroups(seurat_object ~ n_features, features, 
                    method = 2, max.ylev = 6)
t1 <- createTable(c1, show.p.overall = F)
export2md(t1, caption = "")
```




## Apply SCTransform

Apply SCTransform normalization to each subsetted Seurat object in the list.





```{r}
#| eval: false

seurat_objects <- map(seurat_objects, \(x) SCTransform(x))
```

## Integration

### Select integration features

## SCTransform-based integration of multiple Seurat objects

1.  Identifies the top 2,000 highly variable genes across all Seurat objects to use for integration. These features should be common and informative for aligning datasets.

2.  Prepares the `SCT` assays for integration. It calculates the necessary Pearson residuals and scales data based on the selected features. This must be done after SCTransform and before `FindIntegrationAnchors`.


```{r}
#| eval: false
##################
features <- SelectIntegrationFeatures(object.list = seurat_objects, nfeatures = 2000)

# Prepare for integration
seurat_objects <- PrepSCTIntegration(seurat_objects, anchor.features = features)
```

### Find anchors

3.  Identifies anchors—pairs of cells from different datasets that are biologically similar. These are used to align the datasets during integration.



```{r}
#| eval: false
# Find integration anchors
anchors <- FindIntegrationAnchors(
  object.list = seurat_objects,
  normalization.method = "SCT",
  reduction = "cca",
  anchor.features = features
)



```

###  Integrate the datasets

4.  Performs the actual integration, combining all datasets into a single Seurat object where batch effects are minimized. Output is stored in a new `"integrated"` assay.

```{r}
#| eval: false
seurat_integrated <- IntegrateData(
  anchorset = anchors,
  normalization.method = "SCT"
)

save(seurat_integrated,file ="data/20250619-seurat_integrated_nFeat>2000.Rdata")
```

